Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-08-15 manuscript/manuscript.Rmd
1st round revision done based on Dietz's comments.

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology. Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists.

@@

In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights into our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition. After all, these experimental results can only be as robust as their experimental settings.

More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in the realm of philosophy of science [@cartwright2007].

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. Unlike controlled experiments, variability could be easily observed in actualistic experiments by design. This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in recent years [@cunningham2021; @hernan2023]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

@@

Contemporary practices in experimental archaeology, as manifested by the fact that the majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise". The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989]. When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654]. The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists. However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.

@@

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this is not a solid assumption given that 1) experiments and modeling studies in congnitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practioniers in tradional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

@@

Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities. This approach enables researchers to gain a deeper understanding of how people in ancient societies might have perceived and interpreted their environment, objects, and actions. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.

@@

The accessibility and availability of experimental data foster collaboration and enhance the reproducibility and transparency of research findings, as others can verify and validate the results by examining the original data. More importantly, a centralized database also promotes data preservation and long-term accessibility. By storing experimental data in a structured and organized manner, it safeguards valuable information from potential loss or degradation over time. This preservation ensures that the data remains accessible for future researchers, avoiding the loss of valuable insights and preventing the need for redundant and costly repetitions of experiments. It also allows for the reanalysis of existing data, facilitating discoveries and insights that may not have been initially anticipated. However, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].

>>>>

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology. Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].

@@

The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267], and the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition. After all, these experimental results can only be as robust as their experimental settings.

More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. This issue is rarely discussed in material science oriented experiments but should raise a critical concern in experimental archaeology as social science. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in the realm of philosophy of science [@cartwright2007].

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. Unlike controlled experiments, variability could be easily observed in actualistic experiments by design. This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epdedimiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

@@

Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker1999], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise". The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989]. When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654]. The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists. However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.

@@

Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials. It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artifact or a whole assemblage. No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data. Besides, adopting good habits in spreadsheet data organization is strongly recommended [@broman2018].

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

@@

Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.

@@
